---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  eval = FALSE
)
```

# ecotrends

<!-- badges: start -->

<!-- badges: end -->

The goal of `ecotrends` is to **compute a time series of ecological niche models**, using species occurrence data and environmental variables, and then map the existence and direction of **linear temporal trends in environmental suitability**, as in [Arenas-Castro & Sillero (2021)](https://doi.org/10.1016/j.scitotenv.2021.147172).

This package is part of the [MontObEO project](https://montobeo.wordpress.com/).

## Installation

You can install `ecotrends` from GitHub with:

```{r}
devtools::install_github("AMBarbosa/ecotrends")
```

## Usage

You'll need some **species presence coordinates**. The code below downloads some example occurrence data from GBIF, and then performs **just a basic automatic cleaning**:

```{r occurrences}
library(geodata)
# update fuzzySim:
# install.packages("fuzzySim", repos="http://R-Forge.R-project.org")
library(fuzzySim)

occ_raw <- geodata::sp_occurrence(genus = "Chioglossa", 
                                  species = "lusitanica", 
                                  fixnames = FALSE)

occ_clean <- fuzzySim::cleanCoords(data = occ_raw, 
                                   coord.cols = c("decimalLongitude", "decimalLatitude"), 
                                   uncert.col = "coordinateUncertaintyInMeters",
                                   uncert.limit = 10000, 
                                   year.col = "year", 
                                   year.min = 1970, 
                                   abs.col = "occurrenceStatus", 
                                   plot = FALSE)

occ_coords <- occ_clean[ , c("decimalLongitude", "decimalLatitude")]
```

You should also **delimit a region for modelling**. You can provide your own spatial extent or polygon -- e.g., a biogeographical region within which your species was reasonably surveyed. Alternatively or additionally, you can use the `getRegion` function of package `fuzzySim` to compute a "reasonably sized" area around your species occurrences (see help file and try out different options, some of which may be much more adequate for your particular case!):

```{r region}
reg <- fuzzySim::getRegion(pres.coords = occ_coords,
                           CRS = "EPSG:4326",  # make sure it's correct for your case!
                           type = "width",
                           width_mult = 0.5,
                           dist_mult = 1)
```

Now let's **download some variables** with which to build a **yearly time series** of ecological niche models for this species in this region. You can first use the `varsAvailable` function to check which variables and years are available, and then the `getVariables` function to download the ones you choose. Mind that the download may take a long time:

```{r variables}
library(ecotrends)

ecotrends::varsAvailable()

vars <- ecotrends::getVariables(vars = c("tmin", "tmax", "ppt", "pet", "ws"), 
                                years = 1981:1990, 
                                region = reg, 
                                file = "outputs/variables")

names(vars)
plot(vars[[1:6]])
```

The variable raster layers have a given pixel size *at the Equator*, but actual pixel sizes can vary widely across latitudes. So, let's **check the average pixel size in our study region**, as well as the spatial uncertainty values of our occurrence coordinates:

```{r pixelarea}
sqrt(ecotrends::pixelArea(vars))

summary(occ_clean$coordinateUncertaintyInMeters, na.rm = TRUE)
```

You can see there are several occurrence points with spatial uncertainty larger than our pixel size, so it might be a good idea to **coarsen the spatial resolution** of the variable layers:

```{r aggregate}
vars_agg <- terra::aggregate(vars, 
                             fact = 2)

sqrt(ecotrends::pixelArea(vars_agg))
```

This is much closer to the spatial resolution of many of the species occurrences. We can now **compute yearly ecological niche models** with these occurrences and variables, optionally saving the results to a file:

```{r models}
mods <- ecotrends::getModels(occs = occ_coords, 
                             rasts = vars_agg, 
                             region = reg,
                             collin = TRUE, 
                             maxcor = 0.75,
                             maxvif = 5,
                             classes = "default", 
                             regmult = 1, 
                             file = "outputs/models")
```

Let's now **compute the model predictions** for each year, optionally delimiting them to the modelled region, and optionally saving results to a file:

```{r predictions}
preds <- ecotrends::getPredictions(rasts = vars_agg, 
                                   mods = mods$models, 
                                   region = reg,
                                   type = "cloglog",
                                   clamp = TRUE,
                                   file = "outputs/predictions")

plot(preds, range = c(0, 1))
```

You can **evaluate the fit** of these predictions to the model training data:

```{r performance}
perf <- ecotrends::getPerformance(rasts = preds, 
                                  data = mods$data,
                                  plot = TRUE)

perf
```

Finally, you can use the `getTrend` function to **check for a linear (monotonic) temporal trend in suitability** in each pixel, optionally providing your occurrence coordinates if you want the results to be restricted to the pixels that overlap them:

```{r trend}
trend <- ecotrends::getTrend(rasts = preds,
                             occs = occ_coords,
                             alpha = 0.05,
                             full = TRUE,
                             file = "outputs/trend")

plot(trend, 
     col = hcl.colors(100, "spectral"))
```
See `?Kendall::MannKendall` (including the *Value* section) to know more about these statistics. If you want just the first raster layer (with the significant Tau values), set `full = FALSE` above. Positive Tau values indicate increasing suitability, and negative values indicate decreasing suitability over time. Pixels with no value have no significant linear trend (or no occurrence points, if `occs` are provided).
