---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  eval = FALSE
)
```

# ecotrends (version `r packageVersion("ecotrends")`)

<!-- badges: start -->

<!-- badges: end -->

The goal of `ecotrends` is to **compute a time series of ecological niche models**, using species occurrence data and environmental variables, and then map the existence and direction of **linear temporal trends in environmental suitability**, as in [Arenas-Castro & Sillero (2021)](https://doi.org/10.1016/j.scitotenv.2021.147172).

This package is part of the [MontObEO project](https://montobeo.wordpress.com/).

## Installation

You can install `ecotrends` from GitHub with:

```{r}
devtools::install_github("AMBarbosa/ecotrends")
```

## Usage

You'll need some **species presence coordinates**. The code below downloads some example occurrence data from GBIF, and then performs just a **basic automatic cleaning**:

```{r occurrences}
library(geodata)
# installl fuzzySim development version:
# install.packages("fuzzySim", repos="http://R-Forge.R-project.org")
library(fuzzySim)

occ_raw <- geodata::sp_occurrence(genus = "Chioglossa", 
                                  species = "lusitanica", 
                                  fixnames = FALSE)

occ_clean <- fuzzySim::cleanCoords(data = occ_raw, 
                                   coord.cols = c("decimalLongitude", "decimalLatitude"), 
                                   uncert.col = "coordinateUncertaintyInMeters",
                                   uncert.limit = 10000, 
                                   year.col = "year", 
                                   year.min = 1970, 
                                   abs.col = "occurrenceStatus", 
                                   plot = FALSE)

occ_coords <- occ_clean[ , c("decimalLongitude", "decimalLatitude")]
```

You should also **delimit a region for modelling**. You can provide your own spatial extent or polygon -- e.g., a biogeographical region that is **within your species' reach**, and within which that species was **reasonably surveyed** (mind that pixels within your region that don't overlap species presences are taken by Maxent as available and unoccupied). Alternatively or additionally, you can use e.g. the code below to compute a *reasonably sized* area around your species occurrences (see help file and try out different options, some of which may be much more adequate for your particular case!):

```{r region}
reg <- fuzzySim::getRegion(pres.coords = occ_coords,
                           CRS = "EPSG:4326",  # make sure it's correct for your data!
                           type = "width",
                           width_mult = 0.5,
                           dist_mult = 1)
```

Now let's **download some variables** with which to build a **yearly time series** of ecological niche models for this species in this region. You can first use the `varsAvailable` function to check which variables and years are available through the `ecotrends` package, and then the `getVariables` function to download the ones you choose (unless you want to use your own variables from elsewhere). Mind that the download may take a long time:

```{r variables}
library(ecotrends)

ecotrends::varsAvailable()

vars <- ecotrends::getVariables(vars = c("tmin", "tmax", "ppt", "pet", "ws"), 
                                years = 1981:1990, 
                                region = reg, 
                                file = "outputs/variables")

names(vars)
plot(vars[[1:6]])
```

These variable raster layers have a given pixel size in geographic degrees, with a nominal pixel size *at the Equator*, but (as the longitude meridians all converge towards the poles) actual pixel sizes can vary widely across latitudes. So, let's **check the average pixel size in our study region**, as well as the spatial uncertainty values of our occurrence coordinates:

```{r pixelarea}
sqrt(ecotrends::pixelArea(vars))

summary(occ_clean$coordinateUncertaintyInMeters, na.rm = TRUE)
```

You can see there are several occurrence points with spatial uncertainty larger than our pixel size, so it might be a good idea to **coarsen the spatial resolution** of the variable layers:

```{r aggregate}
vars_agg <- terra::aggregate(vars, 
                             fact = 2)

sqrt(ecotrends::pixelArea(vars_agg))
```

This is much closer to the spatial resolution of many of the species occurrences. We can now **compute yearly ecological niche models** with these occurrences and variables, optionally saving the results to a file:

```{r models}
mods <- ecotrends::getModels(occs = occ_coords, 
                             rasts = vars_agg, 
                             region = reg,
                             nbg = 10000,
                             nreps = 3,
                             collin = TRUE, 
                             maxcor = 0.75,
                             maxvif = 5,
                             classes = "default", 
                             regmult = 1, 
                             file = "outputs/models")
```

Note that (if you have `fuzzySim` >= 4.26 installed) you can add a `bias` layer to drive the selection of background points to incorporate survey effort, if your study area contains more pixels than `nbg`. See the `?getModels` help file for more details.

Let's now **compute the model predictions** for each year, optionally delimiting them to the modelled region (though you can predict on a larger or an entirely different region, assuming that the species-environment relationships are the same as in the modelled region), and optionally exporting the results to a file:

```{r predictions}
preds <- ecotrends::getPredictions(rasts = vars_agg, 
                                   mods = mods, 
                                   region = reg,
                                   type = "cloglog",
                                   clamp = TRUE,
                                   file = "outputs/predictions")

names(preds)
plot(preds[[1]], range = c(0, 1))
```

You can **evaluate the fit** of these predictions to the model training data:

```{r performance}
par(mfrow = c(2, 2))

perf <- ecotrends::getPerformance(rasts = preds,
                                  mods = mods,
                                  plot = TRUE)

head(perf)
```

Note that `rasts` here can be either the output of `getPredictions()`, or a `file` argument previously provided to `getPredictions()`, in case you exported predictions in a previous R session and don't want to compute them again.

Finally, you can use the `getTrend` function to **check for a linear (monotonic) temporal trend in suitability** in each pixel (as long as there are more than 3 time steps with suitability values), optionally providing your occurrence coordinates if you want the results to be restricted to the pixels that overlap them:

```{r trend}
trend <- ecotrends::getTrend(rasts = preds,
                             occs = occ_coords,
                             alpha = 0.05,
                             full = TRUE,
                             file = "outputs/trend")

plot(trend, 
     col = hcl.colors(100, "spectral"))
```
See `?Kendall::MannKendall` (including the *Value* section) to know more about these statistics. If you want to compute only the first raster layer (with the significant Tau values), set `full = FALSE` above. Or you can compute the full result as above, but plot just a layer you're interested in, and optionally add the region polygon:

```{r trend2}
plot(trend[["tau"]])
plot(reg, lwd = 0.5, add = TRUE)
```

Positive Tau values indicate increasing suitability, and negative values indicate decreasing suitability over time. Pixels with no value have no significant linear trend (or no occurrence points, if `occs` are provided).
