---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  eval = FALSE
)
```

# ecotrends

<!-- badges: start -->

<!-- badges: end -->

The goal of `ecotrends` is to compute a time series of ecological niche models, using species occurrence data and environmental variables, and then analyse the trend in environmental suitability over time, as in [Arenas-Castro & Sillero (2021)](https://doi.org/10.1016/j.scitotenv.2021.147172).

This package is part of the [MontObEO project](https://montobeo.wordpress.com/).

## Installation

You can install `ecotrends` from GitHub with:

```{r}
devtools::install_github("AMBarbosa/ecotrends")
```

## Users manual

You'll need some species presence coordinates. The code below downloads some example occurrence data from GBIF, and then performs **just a basic automatic cleaning**:

```{r occurrences}
library(geodata)
library(fuzzySim)

occ_raw <- geodata::sp_occurrence(genus = "Chioglossa", species = "lusitanica", fixnames = FALSE)

occ_clean <- fuzzySim::cleanCoords(occ_raw, coord.cols = c("decimalLongitude", "decimalLatitude"), uncert.col = "coordinateUncertaintyInMeters", uncert.limit = 10000, year.col = "year", year.min = 1970, abs.col = "occurrenceStatus", plot = FALSE)

occ_coords <- occ_clean[ , c("decimalLongitude", "decimalLatitude")]
```

You should also delimit a region for modelling. You can provide your own spatial extent or polygon -- e.g., a biogeographical region within which your species was more or less evenly surveyed. Alternatively or additionally, you can use the `getRegion` function to compute a "reasonably sized" area around your species occurrences:

```{r region}
library(ecotrends)

reg <- ecotrends::getRegion(occs = occ_coords)

plot(reg, col = "wheat")
points(occ_coords, cex = 0.3)
```

Now let's download some variables with which to build a yearly time series of ecological niche models for this species in this region. You can first use the `varsAvailable` function to check which variables and years are available, and then the `getVariables` function to download the ones you choose. Mind that the download may take a long time:

```{r variables}
ecotrends::varsAvailable()

vars <- ecotrends::getVariables(vars = c("tmin", "tmax", "ppt", "pet", "ws"), years = 1990:1981, region = reg, file = "variable_rasters")

# or, after you've downloaded the variables with the above 'file' argument:
# vars <- terra::rast("variable_rasters.tif")

names(vars)
plot(vars[[1:6]])
```

The variable raster layers have a given pixel size *at the Equator*, but actual pixel sizes can vary widely across latitudes. So, let's check the average pixel size in our study region, as well as the spatial uncertainty values of our occurrence coordinates:

```{r pixelarea}
sqrt(ecotrends::pixelArea(vars, unit = "m"))

summary(occ_clean$coordinateUncertaintyInMeters, na.rm = TRUE)
```

You can see there are several occurrence points with spatial uncertainty larger than the pixel size, so it might be a good idea to coarsen the spatial resolution of the variable layers:

```{r resolution}
vars_agg <- terra::aggregate(vars, fact = 2)

sqrt(ecotrends::pixelArea(vars_agg, unit = "m"))
```

This is much closer to the spatial resolution of many of the species occurrences. We can now compute yearly models with these occurrences and variables:

```{r models}
mods <- ecotrends::getModels(occ_coords, vars_agg, region = reg, collin = TRUE, file = "models")
```

Let's now compute the model predictions for each year:

```{r preds}
preds <- ecotrends::getPredictions(vars_agg, mods, file = "predictions")

plot(preds)
```

Finally, you can use the `getTrend` function to check for a monotonic temporal trend in suitability in each pixel:

```{r}
trend <- getTrend(preds, alpha = 0.05)

plot(trend, col = hcl.colors(100, "spectral"), main = "Suitability trend")
```

Positive values indicate increasing suitability, and negative values indicate decreasing suitability over time. Pixels with no value have no significant trend.
