---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  eval = FALSE
)
```

# ecotrends

<!-- badges: start -->

<!-- badges: end -->

The goal of `ecotrends` is to **compute a time series of ecological niche models**, using species occurrence data and environmental variables, and then map the existence and direction of **linear temporal trends in environmental suitability**, as in [Arenas-Castro & Sillero (2021)](https://doi.org/10.1016/j.scitotenv.2021.147172).

This package is part of the [MontObEO project](https://montobeo.wordpress.com/).

## Installation

You can install `ecotrends` from GitHub with:

```{r}
devtools::install_github("AMBarbosa/ecotrends")
```

## Usage

You'll need some **species presence coordinates**. The code below downloads some example occurrence data from GBIF, and then performs **just a basic automatic cleaning**:

```{r occurrences}
library(geodata)
library(fuzzySim)

occ_raw <- geodata::sp_occurrence(genus = "Chioglossa", 
                                  species = "lusitanica", 
                                  fixnames = FALSE)

occ_clean <- fuzzySim::cleanCoords(data = occ_raw, 
                                   coord.cols = c("decimalLongitude", "decimalLatitude"), 
                                   uncert.col = "coordinateUncertaintyInMeters",
                                   uncert.limit = 10000, 
                                   year.col = "year", 
                                   year.min = 1970, 
                                   abs.col = "occurrenceStatus", 
                                   plot = FALSE)

occ_coords <- occ_clean[ , c("decimalLongitude", "decimalLatitude")]
```

You should also **delimit a region for modelling**. You can provide your own spatial extent or polygon -- e.g., a biogeographical region within which your species was more or less evenly surveyed. Alternatively or additionally, you can use the `getRegion` function to compute a "reasonably sized" area around your species occurrences:

```{r region}
library(ecotrends)

reg <- ecotrends::getRegion(occs = occ_coords,
                            type = "mean"  # can also be "width"
                            )

plot(reg, col = "wheat")
points(occ_coords, cex = 0.3)
```

Now let's **download some variables** with which to build a **yearly time series** of ecological niche models for this species in this region. You can first use the `varsAvailable` function to check which variables and years are available, and then the `getVariables` function to download the ones you choose. Mind that the download may take a long time:

```{r variables}
ecotrends::varsAvailable()

vars <- ecotrends::getVariables(vars = c("tmin", "tmax", "ppt", "pet", "ws"), 
                                years = 1990:1981, 
                                region = reg, 
                                file = "variable_rasters")

# or, after you've downloaded the variables with the above 'file' argument:
# vars <- terra::rast("variable_rasters.tif")

names(vars)
plot(vars[[1:6]])
```

The variable raster layers have a given pixel size *at the Equator*, but actual pixel sizes can vary widely across latitudes. So, let's **check the average pixel size in our study region**, as well as the spatial uncertainty values of our occurrence coordinates:

```{r pixelarea}
sqrt(ecotrends::pixelArea(vars))

summary(occ_clean$coordinateUncertaintyInMeters, na.rm = TRUE)
```

You can see there are several occurrence points with spatial uncertainty larger than our pixel size, so it might be a good idea to **coarsen the spatial resolution** of the variable layers:

```{r resolution}
vars_agg <- terra::aggregate(vars, fact = 2)

sqrt(ecotrends::pixelArea(vars_agg))
```

This is much closer to the spatial resolution of many of the species occurrences. We can now **compute yearly ecological niche models** with these occurrences and variables, optionally saving the results to a file:

```{r models}
mods <- ecotrends::getModels(occs = occ_coords, 
                             rasts = vars_agg, 
                             region = reg,
                             collin = TRUE, 
                             file = "models")

# or, after you've computed your models with the above 'file' argument:
# mods <- readRDS("models.rds")
```

Let's now **compute the model predictions** for each year, optionally delimiting them to the modelled region, and optionally saving results to a file:

```{r predictions}
preds <- ecotrends::getPredictions(rasts = vars_agg, 
                                   mods = mods, 
                                   region = reg,
                                   file = "predictions")

# or, after you've computed the predictions with the above 'file' argument:
# preds <- terra::rast("predictions.tif")

plot(preds, range = c(0, 1))
```

Finally, you can use the `getTrend` function to **check for a linear (monotonic) temporal trend in suitability** in each pixel, optionally providing your occurrence coordinates if you want the results to be restricted to the pixels that overlap them:

```{r trend}
trend <- ecotrends::getTrend(rasts = preds,
                             occs = occ_coords,
                             alpha = 0.05,
                             full = TRUE)

plot(trend, 
     col = hcl.colors(100, "spectral"))
```
See `?Kendall::MannKendall` (including the 'Values' section) to know more about these statistics. If you want just the Tau value, you can set `full = FALSE` above. Positive Tau values indicate increasing suitability, and negative values indicate decreasing suitability over time. Pixels with no value have no significant linear trend (or no occurrence points, if `occs` are provided).
